---
title: "Data Manipulation"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
knitr::opts_chunk$set(echo=FALSE)
```

##Introduction
In the course of analyzing the bike-sharing data we will come across several problems with the structure of the data. For example we will have to transform/adapt the dates to R classes, or we might want a function to extract a subset of the data. These problems, along with the associated functions that we write will be documented here. 


##Organization of the files
The variable _datapath_ should point to the location of the data within the project directory, and the variable _analpath_ to the location of the analyzes. The data files are zip files under the sub-directories of _d)atapath_. 


##Reading .csv files 
The data is in _.csv_ which may or not may not be present in the folder. If the _.csv_ is not present, it could be in a _.zip_ file. We write a function to take the appropriate action.
```{r loadcsvOrzip}
read.csv.zip <- function(name.csv, path.csv, path.zip){
  if(file.exists(path.csv)){
    read.csv(path.csv, stringsAFactors=FALSE)
  } else {
    if(file.exists(path.zip)){
      read.csv(unz(path.zip, name.csv))
    }
  }
}
```

##Extract data by time periods
```{r sampledata}
library(lubridate)
extractGivenDays <- function(data,
														 start, 
														 end,
														 var.start_time="start_time",
														 var.end_time="end_time"){
	extidx <- (data[ , var.start_time] >= start) & ( data[, var.end_time] <= end)
	data[extidx, ]
}
extractWeek <- function(data, start, 
												var.start_time="start_time",
												var.end_time="end_time",
												weekdays=FALSE){
	extractGivenDays(data, 
	  						   start=start, 
									 end=start + days(ifelse(weekdays, 5, 7)),
									 var.start_time=var.start_time,
									 var.end_time=var.end_time)
}
```


### Transactions and trips
The data provides records for each completed trip. We transform this data to transactions, here each transaction is either a bike rent or a bike return. Thus each trip will be converted into two transactions, one going _out_ (rent) and one _in_ (return).  

```{r makeTransaction}
makeTrxn <- function(trips){
	n <- nrow(trips)
  allvars <- names(trips)
  tocopy <- setdiff(allvars, c("start_time", "end_time", 
                               "start_statn", "end_statn"))
  df <- rbind( trips[, tocopy], trips[, tocopy])
	cbind(df, with(trips, data.frame(time=c(start_time, end_time),
                    			         statn=c(start_statn, end_statn),
                    			         drxn=c(rep("out", n), rep("in", n))
                                  )
                 )
        )
}
```
