---
title: "Hubway Bike Sharing"
output: html_document
---
Whats in the Hubway bike sharing data?
```{r loadData}
hubway.orig <- read.csv("data/hubway/hubway_trips.csv", stringsAsFactor=FALSE)
hubway <- hubway.orig
hubway.st <- read.csv("data/hubway//hubway_stations.csv")
```
Lets use a datatime class for the dates,
```{r datetimefix}
library(lubridate)
hubway$start_date <- mdy_hms(hubway$start_date)
hubway$end_date <- mdy_hms(hubway$end_date)
```

There are over 1 million records in the hubway data-set. This will make our examples slow to run, so we take a sample of the available data to use,
```{r sampledata}
N <- 10000
extractGivenDays <- function(data=hubway, start, end){
  subset(hubway, start_date >= start & end_date <= end)
}
extractWeek <- function(data=hubway, start, weekdays=FALSE){
  extractGivenDays(data, start=start, end=start + days(ifelse(weekdays, 5, 7)) )
}
```
We find that the first 35000 rows give us about one month of data, which is a good number to start with. After we have trained some models, we can look back at the full data-set

##Number of bikes at a station.
Let us count the number of bikes going in minus the number going out of a station, as a function of time. For this we need to convert data time to datetime classes,
Each record in the data is a full trip. We need to convert this to transactions. 
```{r tripsToTransactions}
#refactor this
#there should be a faster way of converting the trips to transactions, but this was the
#simplest implementation
makeTransaction <- function(trip){
  trans <- data.frame(
    type=c("out", "in"),
    bike=c(trip$bike_nr, trip$bike_nr),
    time=c(trip$start_date, trip$end_data),
    loc=c(trip$strt_statn, trip$end_statn)
  )
  trans$bike <- as.character(trans$bike)
  trans
}
hubway.sample <- extractWeek(data=hubway, 
                             start=ymd_hms("2012-06-04 0:0:0"), 
                             weekdays=TRUE)
                                  
hubway_trans <- do.call(rbind, 
                        lapply(1:nrow(hubway.sample), 
                               function(i) makeTransaction(hubway[i,])))
```
We ran the code above, and it takes too long, and we terminated it. 

A better thing would be to extract a time series for any station from the trips data.
```{r stationTS}
station_timeseries <- function(st, data=hubway.sample) {
  st.bgn <- subset(data, strt_statn == st)
  nb <- nrow(st.bgn)
  if( nb > 0){
    st.bgn$time <- st.bgn$start_date
    st.bgn$dcount <- -1
    st.bgn$dir <- "out"
  }
  st.end <- subset(data, end_statn == st)
  ne <- nrow(st.end)
  if(ne > 0){
    st.end$time <- st.end$end_date
    st.end$dcount <- 1
    st.end$dir <- "in"
  }
  st.all <- rbind(st.bgn, st.end)
  if(nrow(st.all) > 0 ){
    st.all <- st.all[ order(st.all$time), ]
  }
  st.all$count <- cumsum(st.all$dcount)
  st.all$count <- st.all$count + ifelse(st.all$count[1] == 1, -1, 1)
  st.all$volume <- 1:nrow(st.all)
  if(nrow(st.all) > 0) {
    with(data=st.all, data.frame(station=st, time=time, count=count, volume=volume,dir=dir))
  }
  else {
    data.frame(station=st, time=NA, count=NA, volume=NA, dir=NA)
  }
}
```
Lets look at some of these time-series. We need to know what station ids are in the data set.
```{r timeseries}
stns <- unique(c(hubway.sample$strt_statn, hubway.sample$end_statn))
stns <- stns[!is.na(stns)]
plot(count ~ time, data=station_timeseries(data=hubway.sample, 11),type="b")
plot(volume ~ time, data=station_timeseries(data=hubway.sample, 11),type="b")
timeseries.st <- do.call(rbind, lapply( stns, function(s){
  station_timeseries(st=s, data=hubway.sample)
}) )
timeseries.st <- subset(timeseries.st, time < max(hubway.sample$start_date))
```
Make plots 
```{r tsplots}
library(ggplot2)
ggplot(data=subset(timeseries.st, station %in% stns[1:10]),
       aes(x=time, y=count, col=factor(station))) + 
  geom_line()
```
###Symmetry between stations
Some functions that define the symmetry of bike flow between pairs of stations,
```{r stsym}
#flow from station st1 to st2
flowBwStns <- function(data=hubway.sample, st1, st2){
  hbw <- subset(hubway.sample, 
                (strt_statn == st1 & end_statn == st2) | (strt_statn == st2 & end_statn == st1) )
  fl <- cumsum(2*(hbw$strt_statn == st1) - 1)
  hbw$flow <- fl + ifelse(fl[1] == -1, 1, -1) #start the flow from 0
  with(hbw, data.frame(from=st1, to=st2, start_time=start_date, end_time=end_date, flow=flow))
}

#net flow between stations at the end of a day
balanceBwStns <- function(data=hubway.sample, st1, st2){
  fb <- flowBwSts(data, st1, st2)
  alldays <- unique(day(fb$end_time))
  do.call("rbind", lapply(alldays, function(d){
      fbd <- subset(fb, day(end_time) == d)
      fbd <- fbd[ order(fbd$end_time), ]
      with(fbd[ nrow(fbd), ], data.frame(from=st1, to=st2, day=d, flow=flow))
    })
  )
}
balanceDayEnd <- function( st, st2 = NULL, data=hubway.sample){
  last <- function(dayData){
    dayData <- dayData[ order(dayData$time),]
    dayData$count[ nrow(dayData)]
  }
  getDay <- function(date) {
    paste(year(date), month(date), day(date), sep="-")
  }
  if (is.null(st2)){
    stdata <- station_timeseries(st)
    alldays <- unique(getDay(stdata$time))
    do.call("rbind", lapply(alldays, function(d){
      dayData <- subset(stdata, getDay(time) == d)
      data.frame(statn=st, day=ymd(d), balance=last( dayData))
    }))
  }
  else{
    stdata <- flowBwStns(data, st, st2)
    stdata$time <- stdata$start_time
    alldays <- unique(getDay(fb$end_time))
    do.call("rbind", lapply(alldays, function(d){
      dayData <- subset(stdata, getDay(time) == d)
      data.frame(strt_statn=st, end_statn=st2, balance=last(dayData))
    }))
  }
}
  




