---
title: "Bike Sharing User Perspective"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
knitr::opts_chunk$set(echo=FALSE)
```
##Abstract
What does the user think about the bike sharing system they are using. 

##Data and Exploratory Analysis

The example data that we use is the hubway data, 
```{r loadData}
hubway.orig <- read.csv("data/hubway/hubway_trips.csv", stringsAsFactor=FALSE)
hubway <- hubway.orig
hubway.st <- read.csv("data/hubway//hubway_stations.csv")
```
The hubway data contains columns,
```{r hubwayClms}
names(hubway)
```
We redefine the dates to be useful datetime classes,
```{r datetimefix}
library(lubridate)
hubway$start_date <- mdy_hms(hubway$start_date)
hubway$end_date <- mdy_hms(hubway$end_date)
```

The hubway data is large. In order to run our analysis in reasonable time, and also to focus on given time periods we will sample a subset specified by days,

```{r sampledata}
N <- 10000
extractGivenDays <- function(data=hubway, start, end){
  subset(hubway, start_date >= start & end_date <= end)
}
extractWeek <- function(data=hubway, start, weekdays=FALSE){
  extractGivenDays(data, start=start, end=start + days(ifelse(weekdays, 5, 7)) )
}
```

As an example for our analysis we will sample one week from June 2012.
```{r sampleWeek}
hubway.sample <- extractWeek(data=hubway, 
                             start=ymd_hms("2012-06-04 0:0:0"), 
                             weekdays=TRUE)
```
                                  

We find that the first 35000 rows give us about one month of data, which is a good number to start with. After we have trained some models, we can look back at the full data-set.

### Adding features to the data

Biker behavior in the morning can be different from that in the afternoon. We will divide each day into periods,
```{r dayperiods}
dayPeriods <- data.frame(
  period=c("morn.early", "morn", "aftn", "aftn.late", "evn", "nite", "nite.late"),
  start=c(5, 8, 12, 15, 18, 21, 0),
  end=c(8,12,15,18,21,24,5)
)
dayPeriods$period <- as.character(dayPeriods$period)
st <- hour(hubway.sample$start_date)
hubway.sample$period <- sapply(st, function(s){
  with(dayPeriods, period[ start <= s & end > s])
})
```  

###Disambiguating User Ids

We expect our case-studies to include a user-id. These are not available for the hubway data, however some user data is provided. We discuss various methods in which we can artificially create *user_ids* for the hubway data. A valid user cannot occupy two bikes at the same time.
```{r validuser}
validUser <- function(uid, data=hubway.sample){
  hs <- subset(data, user_id == uid)
  hs <- subset(hs, !is.na(start_date) & !is.na(end_date) )
  hs <- hs[order(hs$start_date),]
  n <- nrow(hs)
  s <- hs$start_date
  e <- hs$end_date
  ifelse(n==1, s <= e, all(s[2:n] > e[1:(n-1)]))
}

validUserIds <- function(uids=NULL, 
                         user_id_var = NULL, 
                         data=hubway.sample){
  if(is.null(uids)) uids <- unique(data[, user_id_var])
  all( sapply(uids, function(u) validUser(u, data)))
}
```
Two trips overlap if their time-lines do.
```{r tripoverlap}
ovlpngTrips <- function(trip1, trip2, 
                        var.start_time="start_date",
                        var.end_time="end_date"){
  s1 <- trip1[, var.start_time]; e1 <- trip1[, var.end_time] 
  s2 <- trip2[, var.start_time]; e2 <- trip2[, var.end_time] 
  if ((s2 > e2) & (s1 > e1)) NA
  else{
    nonovlp <- (e2 <= s1) | (s2 >= e1)
    !nonovlp
  }
}

tripOverlaps <- function(trips, 
                        var.end_time="end_date", 
                        var.start_time="start_date",
                        rentry.time = 0) {
  #use a exponential relaxation to re-enter users into the system after a trip.
  #parameter rentry.time is measured in minutes
  n <- nrow(trips)
  as.dist( do.call("rbind", 
                   lapply(1:n, function(i){
                    dt <- trips[1:i, var.end_time] - trips[i, var.start_time]
                    dt <- as.double(dt, units="mins")
                    dis <- c((dt > 0)*(1 - exp(-dt/rentry.time)),  rep(NA, n - i) ) 
                    dis[is.nan(dis)] <- 0
                    dis
                   }))
        )
}
```

Things are fine, we can use clustering to disambiguate the users. We do this by first defining *pre-user-id*- to be gender+birth-date,
```{r disambg}
hubway.sample$pre_user_id <- with(hubway.sample, paste(gender, birth_date, sep=""))
usrs <- unique(hubway.sample$pre_user_id)

disambguser <- function(user,
                        pre_user_id_var="pre_user_id",
                        var.end_time="end_date",
                        var.start_time="start_date",
                        rentry.time=0,
                        trips=hubway.sample){
  trips <- trips[ trips[, pre_user_id_var] == user,]
  tol <- tripOverlaps(trips, var.end_time, var.start_time, rentry.time)
  trips$new_user_id <- paste(trips[, pre_user_id_var],
                             if( sum(tol) > 0 ) cutree(hclust(tol,method="ward.D2"), h=0) 
                              else   rep(1, nrow(trips)),
                             sep="."
                       )
  trips
}

hubway.sample <- do.call("rbind", lapply(unique(hubway.sample$pre_user_id), 
                                function(u) disambguser(user=u, trips=hubway.sample)))
```
Are we sure that *disambguser* works? The resulting *new_user_id*s should be valid. We will check if each newly generated user's trips overlap.
```{r testdisambg}
userTripsOverlaps <- function(user, trips=hubway.sample, var.user_id="new_user_id"){
  trips.u <- trips[ trips[, var.user_id] == user, ]
  n <- nrow(trips.u)
  if (n > 1) {
    any(sapply(1:(n-1), function(i) {
        #print(paste( "testing trip ", i))
        any(sapply((i+1):n, function(j){
        #  print(paste("\t against trip", j))
          ovlpngTrips(trips.u[i,], trips.u[j,])
        })) 
      }
    ))
  }
  else FALSE
}

newUsers <- unique(hubway.sample$new_user_id)
any( sapply(newUsers, function(u) {
#      print(paste("testing user", u))
      userTripsOverlaps(user=u)
    })
  )
```
The test passed. So now we can establish the new user ids,
```{r hwuid}
hubway.sample$user_id <- hubway.sample$new_user_id
```

To generalize this, we will write a few tests here, just for documentation. To run the tests we will have to export the R-code to a source file (.R) and make a test-directory etc.

#### Tests
```{r testthatdisambg}
#library(testthat)
#test_that("Disambiguates to a single user", {
 # trips <- data.frame(pre_user_id = rep(1, 11), 
  #                    start_date=seq(0, 10, length=11),
   #                   end_date=seq(0, 10, length=11) + 0.5 )
  #trips.da <- disambguser(user=1, trips=trips)
#  expect_that(length(unique(trips.da$new_user_id)), equals(1))
#})
#test_that("Disambiguated users's trips do not overlap", {
 # trips <- data.frame(pre_user_id = rep(1, 11), 
  #                    start_date=seq(0, 10, length=11),
  #                    end_date=seq(0, 10, length=11) + 1.5 )
  #trips.da <- disambguser(user=1, trips=trips)
  #users <- trips.da$new_user_id
  #expect_that(any( sapply(users, 
   #                       function(u) userTripsOverlaps(user=u, trips=trips.da))
    #          ), 
     #         equals(FALSE)
      #       )
#}
```

```{r tabuser}
library(lattice)
tb <- table(hubway.sample$new_user_id)
tb <- data.frame(name=names(tb), freq.trips=as.numeric(tb))
histogram(tb$freq.trips)
```
The histogram look reasonable, so we can go ahead and use our disambiguation method.

### Transactions
The data provides records for each completed trip. We transform this data to transactions, here each transaction is either a bike rent or a bike return. Thus each trip will be converted into two transactions, one going _out_ (rent) and one _in_ (return).  

```{r makeTransaction}
makeTransaction <- function(trips){
    trxns <- with(trips, data.frame(
      statn_loc=c(strt_statn, end_statn),
      user_id=rep(user_id,2),
      trxn_type=c(rep("out", nrow(trips)), rep("in", nrow(trips))),
      bike_nr=rep(bike_nr, 2),
      time=c(start_date, end_date),
      subsc_type=rep(subsc_type,2),
      birth_date=rep(birth_date,2),
      gender=rep(gender,2) )
    )
    trxns$bike_nr <- as.character(trxns$bike_nr)
    trxns
}
```
The columns in the transactions are,
```{r transclmns}
names(makeTransaction(hubway.sample[1,]))
```
The column *statn_loc* should provide the location of the bike-station, either directly as (long, lat) or as a label along with a dictionary that can be queried with that label.
The hubway data has birth-dates and gender for the users, which we have included here, however we may not want to use this data if it will unavailable in our case-studies.

```{r hwmaketrans}
hubway_trans <- makeTransaction(hubway.sample)
hubway_trans$period <- sapply(hour(hubway_trans$time), 
                              function(t) with(dayPeriods, 
                                               period[ t>=start & t<end]) )
```

We can take a peek at the transactions in our sample week,
```{r transpeek}
head(hubway_trans)
```

In some of our analysis we will need the data to be in the form of trips. We will need functions to make translate from transactions to trips, however for the moment we will just use *hubway.sample* data set for this purpose.
```{r trxntotripstub}
#this function is a stub
makeTrip <- function(trxns, 
                     var.start_statn="strt_statn",
                     var.end_statn="end_statn",
                     var.user_id="user_id"
                     ){
  hubway.sample
}
```


##User features
In order to study the users, we extract features from their usage data. The features we will discuss are as follows,
1. Frequency of use
2. Duration of trips
3. Intermittency of use
4. The diversity of start stations
5. The diversity of end stations
6. The diversity of (start, end) station pairs.

To illustrate the features we will use the *hubway_trans* data with disambiguated user ids, as our starting data. We will need trips for some of our computations. 
```{r hwuidftr}
hubway_trips <- makeTrip(trxns=hubway_trans)
users <- unique(hubway_trans$user_id)
```
### Frequency of use
A user's frequency is the number of times they have used the bike-sharing system, or the number of trips they have made. We will assume that a user who rents a bike also returns it in the dataset.
```{r userfreq}
library(reshape2)
library(plyr)
user.frequencies <- function(trxns, 
                      var.trxn_type="trxn_type", 
                      var.user_id="user_id"){
  trxn.types <- unique(trxns[, var.trxn_type])
  users <- unique(trxns[, var.user_id])
  tb <- as.data.frame(table(trxns[, var.user_id], trxns[, var.trxn_type]))
  names(tb) <- c("user_id", "drxn", "freq")
  dcast(data=tb, formula=user_id ~ drxn, value.var="freq", fun.aggregate=sum)
}

ufreqs <- user.frequencies(trxns=hubway_trans)
```

Lets explore user frequencies,
```{r}
ufreqs.l <- melt(ufreqs, id.vars = "user_id", 
                 variable.name = "drxn", value.name="count" )
histogram(~ count | drxn, data=ufreqs.l)
```

### Frequency of use by time of day
We have the variable _period_ that we can use to study the period of the day when the bikes are used. 
```{r periodbars}
prdUsg <- with(hubway_trans, as.data.frame(table(gender, period, trxn_type)))
levels(prdUsg$period) <- c("morn.early", "morn", 
                          "aftn", "aftn.late", 
                          "evn", "evn.late", 
                          "nite", "nite.late")

gp <- ggplot(data=prdUsg, aes(x=period, y=Freq, fill=gender)) +
        geom_bar(stat="identity", position="dodge") + 
          facet_grid( trxn_type ~ . ) + 
            ggtitle("Usage going out by period of day across genders")
print(gp)

gp.out <- ggplot(data=subset(prdUsg, trxn_type=="out"),
                             aes(x=period, y=Freq, fill=gender)) +
            geom_bar(stat="identity", position="dodge") + 
              ggtitle("Usage going out by period of day across genders")
print(gp.out)

gp.in <- ggplot(data=subset(prdUsg, trxn_type=="in"),
                             aes(x=period, y=Freq, fill=gender)) +
            geom_bar(stat="identity", position="dodge") +
              ggtitle("Usage coming in by period of day across genders")
print(gp.in)
```
We see that usage by both males and females follows the same pattern during the day. We can compute the frequency of use across the day for each user,
```{r userdayperiod}
user.prdusg <- function(trxns, 
                        prds,
                        var.user_id="user_id",
                        var.trxn_type="trxn_type",
                        var.time="time",
                        var.period="period"
                        ){
  users <- unique(trxns[, var.user_id])
  #prds <- unique( trxns[, var.period])
  df <- do.call(rbind, lapply(users, function(u){
          tb <- table(trxns[ trxns[, var.user_id] == u, var.period])
          nu <- rep(0, length(prds))
          names(nu) <- prds
          nu[names(tb)] <- tb
          nu
        }))
  df <- data.frame(users=users, df, stringsAsFactors=FALSE)
  names(df)[1] <- var.user_id
  df
}
#eg usage
prds <- c("morn.early", "morn", "aftn", "aftn.late", "evn", "nite", "nite.late")
updu <- user.prdusg(trxns, prds)
```

###Intermittency of use
```{r userinter}
intermittencyOfUser <- function(user, trips,
                               var.user_id="user_id",
                               var.start_time="start_date",
                               var.end_time="end_date"){
  #print(paste("computing intermittency for user", user))
  trips.u <- trips[ trips[, var.user_id] == user,]
  trips.u <- trips.u[order(trips.u[, var.start_time]),]
  start.u <- trips.u[ , var.start_time]
  end.u <- trips.u[, var.end_time]
  n <- nrow(trips.u)
  if(n > 1) difftime(start.u[2:n], end.u[1:(n-1)], units="secs")
  else c()
}
                                
user.intermittencies <- function(trips,
                               var.user_id="user_id",
                               var.start_time="start_date",
                               var.end_time="end_date",
                               statistics=TRUE,
                               .funcs=c(mean, var),
                               .funcs.names=c("mean", "var")){
  #.func.names needs some improvement, as now it wont generalize
  users <- unique(trips[, var.user_id])
  uintmt <- lapply(users, function(u) {
    intermittencyOfUser(u, trips,
                        var.user_id,
                        var.start_time,
                        var.end_time)
            })
  if (statistics){
    df <- as.data.frame(do.call(rbind, lapply(uintmt, function(u) {
        if(length(u) == 0) rep(NA, length(.funcs))
        else sapply(.funcs, function(f) f(u))
      }))
    )
    df <- data.frame(user_id=users, df)
    names(df)[2:3] <- .funcs.names
    df
  }
  else uintmt
}
  
uintmtc <- user.intermittencies(trips=hubway_trips)  
```  
###Duration of trips
Duration of trips can be easily extracted from *trips*
```{r userdurs}
user.durations <- function(trips,
                           var.user_id="user_id",
                           var.start_time="start_date",
                           var.end_time="end_date",
                           statistics=TRUE,
                           .funcs=c(mean, var),
                           .funcs.names=c("mean", "var")){
  #.func.names needs some improvement, as now it wont generalize
  users <- unique(trips[, var.user_id])
  udurs <- lapply(users, function(u) {
    trips.u <- trips[ trips[, var.user_id] == u,]
    difftime(trips.u[ , var.end_time], trips.u[, var.start_time], units="secs")
  })
  if (statistics){
    df <- as.data.frame(do.call(rbind, lapply(udurs, function(u) {
        if(length(u) == 0) rep(NA, length(.funcs))
        else sapply(.funcs, function(f) f(u))
      }))
    )
    df <- data.frame(user_id=users, df)
    names(df)[2:3] <- .funcs.names
    df
  }
  else udurs
}
  
udurs <- user.durations(trips=hubway_trips)  
```  
###Diversity of stations
There aere three diversity values we can define for the stations that a user frequents, start, end and the start-to-end stations of each trip.
```{r userstartstatn}
#remove this line
trxns <- hubway_trans
entropy <- function(ps){
  ps <- ps/sum(ps)
  ps <- ps[ps > 0]
  -sum(ps*log2(ps))
}
user.subset <- function(user, data=hubway_trxns, var.user_id="user_id"){
  data[data[ , var.user_id] == user,]
}
user.stnentropies <- function(trxns,
                             var.user_id="user_id",
                             var.trxn_type="trxn_type",
                             var.statn_id="statn_loc"){
  users <- unique(trxns[, var.user_id])
  data.frame(user_id=users, 
             strtent = sapply(users, function(u){
               trxns.u <- user.subset(u, data=trxns)
               ss <- trxns.u[ trxns.u[, var.trxn_type] == "out", var.statn_id]
               entropy(table(ss))
             }),
             endent = sapply(users, function(u){
               trxns.u <- user.subset(u, data=trxns)
               es <- trxns.u[ trxns.u[, var.trxn_type] == "in", var.statn_id]
               entropy(table(es))
             }),
             strtend = sapply(users, function(u){
               trxns.u <- user.subset(u, data=trxns)
               ss <- trxns.u[ trxns.u[, var.trxn_type] == "out", var.statn_id]
               es <- trxns.u[ trxns.u[, var.trxn_type] == "in", var.statn_id]
               ses <- paste(ss, es, sep=".")
               entropy(table(ses))
             })
  )
}
  
uents <- user.stnentropies(trxns=trxns)  
```


### Gathered user features
Ok, now lets put all that we have together,
```{r featureDataSet}
user.features <- merge(ufreqs, uintmtc, by="user_id")
user.features <- merge(user.features, udurs, by="user_id")
user.features <- merge(user.features, uents, by="user_id")
```
We can now cluster the users,
```{r usrclus}
user.ftr.sc <- scale(user.features[,-1])
user.ftr.sc <- data.frame(user_id=user.features$user_id, user.ftr.sc)
user.ftr.dis <- dist(user.ftr.sc[,-1], method="euclidean")
user.ftr.mat <- as.matrix(user.ftr.sc[,-1])
hc <- hclust(user.ftr.dis , method="ward.D2")
plot(hc)
heatmap(user.ftr.mat)
library(ggplot2)
user.ftr.sc <- user.ftr.sc[ order(cutree(hc, h=20)),]
ufm.m <- melt(user.ftr.sc[, -2], id.vars = "user_id")
p <- ggplot(ufm.m, aes(variable, user_id)) + 
      geom_tile(aes(fill=sign(value)*log10(abs(value))), color="white") +
       scale_fill_gradient(low="yellow", high="red")
print(p)
```
