---
title: "Bike Sharing from a user's perspective"
output:
  pdf_document: default
  html_document:
    highlight: tango
    number_sections: yes
    theme: journal
---
```{r setup, include=TRUE}
knitr::opts_chunk$set(cache=FALSE)
knitr::opts_chunk$set(echo=TRUE)
knitr::opts_chunk$set(fig.width=14)
knitr::opts_chunk$set(fig.height=14)
prjpath <- "~/work/citiviz/bikeSharing/bikestats/"
datapath <- paste(prjpath, "data/", sep="")
analpath <- paste(prjpath, "analyses/", sep="")
rcodepath <- paste(analpath, "Rcode/", sep="")
setwd(analpath)
```
#Abstract
What does the user think about the bike sharing system they are using. 

#Organization of the files
The variable _datapath_ should point to the location of the data within the project directory, and the variable _analpath_ to the location of the analyzes. The data files are zip files under the sub-directories of _datapath_. 

#Data and Exploratory Analysis

For our examples we will use the hubway data, that we load using our custom function,
```{r loadData, cache=TRUE}
source(paste(rcodepath,"datamanip.R", sep=""))
path.zip <- paste(datapath, "hubway/hubway_2011_07_through_2013_11.zip", sep="")
print(paste("we will try to read the file from", path.zip))
print(paste("see if file exists", file.exists(path.zip))) 
hubway.orig <- read.csv.zip(name.csv="hubway_trips.csv", 
                            path.csv=paste(datapath, 
					   "hubway/hubway_trips.csv", sep=""),
                            path.zip=path.zip )


hubway <- hubway.orig
hubway.st <- read.csv.zip(name.csv="hubway_stations.csv",
                          path.csv=paste(datapath, 
					 "hubway/hubway_stations.csv", sep=""),
                          path.zip=path.zip)
print("data read with dimensions")
print(dim(hubway))

```
The hubway data contains columns,
```{r hubwayClms, echo=TRUE}
print(names(hubway))
```
One problem is using different data-sets is the non-uniformity of variable names. We have written a function, _namingScheme_ (see _varNamingScheme.Rmd_) that can provide a dictionary to rename the relevant variables. We illustrate its use in the code below,
```{r renamevars, echo=TRUE, cache=FALSE}
source(paste(rcodepath, "varNamingScheme.R", sep=""))
nms.indata <- names(hubway)
nms.touse <- namingScheme(nms.indata,
                          var.user_id = NA,
                          var.start_statn="strt_statn",
                          var.end_statn="end_statn",
                          var.start_time="start_date",
                          var.end_time="end_date",
                          var.bike_id="bike_nr",
                          var.trip_id=NA)
print("we will rename data from")
print(nms.indata)
print("to")
print(nms.touse)
```

We redefine the dates to be useful datetime classes,
```{r datetimefix, cache=FALSE}
library(lubridate)
newnames <- nms.touse[names(hubway)]
names( hubway ) <- newnames
hubway$start_time <- mdy_hms(hubway$start_time)
hubway$end_time <- mdy_hms(hubway$end_time)
print("the names of hubway should have changed to ")
print(names(hubway))
```

The hubway data is large. In order to run our analysis in reasonable time, and also to focus on given time periods we will sample a subset specified by days. As an example for our analysis we will sample one week from June 2012.
```{r sampleWeek}
print("we will extract data from dims")
print(dim(hubway))
source(paste(rcodepath,"datamanip.R", sep=""))
data.sample <- extractWeek(data=hubway, 
                             start=ymd_hms("2012-06-04 0:0:0"),
			     var.start_time="start_time",
			     var.end_time="end_time",
                             weekdays=TRUE)
print(paste("Number of trips during the week starting on June 04 2012", 
            nrow(data.sample)))
```
                                  
We will use this sample data to develop our statistical analyses. After we have trained some models, we can look back at the full data-set.


# Adding features to the data

Biker behavior in the morning can be different from that in the afternoon. We will divide each day into periods,
```{r dayperiods}
periodOfDay <- function(times){
  dayPeriods <- data.frame(
    period=c("morn.early", "morn", "aftn", "aftn.late", "evn", "nite", "nite.late"),
    start=c(5, 8, 12, 15, 18, 21, 0),
    end=c(8,12,15,18,21,24,5)
  )
  dayPeriods$period <- as.character(dayPeriods$period)
  if (class(times) != "integer") times <- hour(times)
  sapply(times, function(t) {
    with(dayPeriods, period[start <= t & end > t])
  })
}

data.sample$period <- periodOfDay(data.sample$start_time)
```  

#Inventing Users
The hubway data does not provide user ids, which limits the analyses for user properties. To make this data more useful we have written some functions to invent users in 
_userDisambigation.Rmd_.

```{r disambg}
source(paste(rcodepath, "userDisambiguation.R", sep=""))
data.sample$pre_user_id <- with(data.sample, paste(gender, birth_date, sep=""))
usrs <- unique(data.sample$pre_user_id)

data.sample <- do.call("rbind", lapply(unique(data.sample$pre_user_id), 
                                function(u) disambguser(user=u, trips=data.sample)))

newUsers <- unique(data.sample$new_user_id)
any( sapply(newUsers, function(u) {
#      print(paste("testing user", u))
      doUserTripsOverlap(user=u, data.sample, test_ids = "new_user_id")
    })
)
```
The test passed. So now we can establish the new user ids,
```{r hwuid}
data.sample$user_id <- data.sample$new_user_id
```

```{r tabuser}
library(lattice)
tb <- table(data.sample$user_id)
tb <- data.frame(name=names(tb), freq.trips=as.numeric(tb))
histogram(tb$freq.trips, main="Histogram of number of trips made by a users")
```
The histogram look reasonable, so we can go ahead and use our disambiguation method to create user ids.

#Transactions
We have written a function to convert trips to transactions (see _datamanip.Rmd_), that we use to convert the trips to transactions in the hubway sample.

The columns in the transactions are,
```{r transclmns}
names(makeTrxn(data.sample[1,]))
```
The column *statn* should provide the location of the bike-station, either directly as (long, lat) or as a label along with a dictionary that can be queried with that label.
From now on we will have two forms for the data, as trips, and as transactions.
```{r hwmaketrans}
trips.sample <- data.sample
trxns.sample <- makeTrxn(trips.sample)
trxns.sample$period <- periodOfDay( trxns.sample$time)
```

We can take a peek at the transactions in our sample week,
```{r transpeek}
head(trxns.sample)
```

Transalting transactions to trips is harder, and will be implemented later.

#Statistics for a single user
How often does a user use bikes? At what time of day? Which stations are most frequented by a user? What is the duration of a user's trips? 

```{r usagetrace}
usagetrace <- function(user, trips, dt=minutes(5)){
	userTrips <- subset(trips, user_id == user)
  #make a sequence by 5 minutes
  t0 <- min(userTrips$start_time)
  tl <- max(userTrips$end_time)
  N <- duration(interval(start=t0, end=tl))/as.duration(dt)
  tvec <- t0 + (0:N)*dt 
  tripIndices <- lapply(1:nrow(trips), function(i) {
    which( tvec > userTrips$start_time[i] & tvec < userTrips$end_time[i])
  })
  tripIndices <- do.call(c, tripIndices)
  inuse <- rep(0, length(tvec))
  inuse[tripIndices] <- 1
  data.frame(time=tvec, inuse=inuse, user_id=user)
}
```
Lets explore usage traces of some users,
```{r explusgtrcs}
allusers <- unique(trips.sample$user_id)
eg.ustrs <- do.call(rbind, lapply(allusers[1:16], function(u) usagetrace(u, trips.sample)))
xyplot( inuse ~ time | user_id, data=eg.ustrs, type="l")
```

What about the intermittency of use by an individual user? 

#User features

In order to study the users, we extract features from their usage data. The features we will discuss are as follows,
1. Frequency of use
2. Duration of trips
3. Intermittency of use
4. The diversity of start stations
5. The diversity of end stations
6. The diversity of (start, end) station pairs.

To illustrate the features we will use both the trip and transaction forms of the data with disambiguated user ids, as our starting data. 
### Frequency of use
A user's frequency is the number of times they have used the bike-sharing system, or the number of trips they have made. We will assume that a user who rents a bike also returns it in the dataset.
```{r userfreq}
library(reshape2)
library(plyr)
user.frequencies <- function(trxns){
  trxn.types <- unique(trxns$drxn )
  users <- unique(trxns$user_id)
  tb <- as.data.frame(table(trxns$user_id, trxns$drxn))
  names(tb) <- c("user_id", "drxn", "freq")
  dcast(data=tb, formula=user_id ~ drxn, value.var="freq", fun.aggregate=sum)
}

ufreqs <- user.frequencies(trxns=trxns.sample)
```

Lets explore user frequencies,
```{r}
ufreqs.l <- melt(ufreqs, id.vars = "user_id", 
                 variable.name = "drxn", value.name="count" )
histogram(~ count | drxn, data=ufreqs.l)
```

## Frequency of use by time of day
We have the variable _period_ that we can use to study the period of the day when the bikes are used. 
```{r periodbars}
library(ggplot2)
print(names(trxns.sample))
prdUsg <- with(trxns.sample, as.data.frame(table(gender, period, drxn)))
levels(prdUsg$period) <- c("morn.early", "morn", 
                          "aftn", "aftn.late", 
                          "evn", "evn.late", 
                          "nite", "nite.late")

gp <- ggplot(data=prdUsg, aes(x=period, y=Freq, fill=gender)) +
        geom_bar(stat="identity", position="dodge") + 
          facet_grid( drxn ~ . ) + 
            ggtitle("Usage going out by period of day across genders")
print(gp)

gp.out <- ggplot(data=subset(prdUsg, drxn =="out"),
                             aes(x=period, y=Freq, fill=gender)) +
            geom_bar(stat="identity", position="dodge") + 
              ggtitle("Usage going out by period of day across genders")
print(gp.out)

gp.in <- ggplot(data=subset(prdUsg, drxn =="in"),
                             aes(x=period, y=Freq, fill=gender)) +
            geom_bar(stat="identity", position="dodge") +
              ggtitle("Usage coming in by period of day across genders")
print(gp.in)
```
We see that usage by both males and females follows the same pattern during the day. We can compute the frequency of use across the day for each user,
```{r userdayperiod}
user.prdusg <- function(data, 
                        prds){
  users <- unique(data$user_id)
  df <- do.call(rbind, lapply(users, function(u){
					tb <- table(subset(data, user_id == u)$period)
          nu <- rep(0, length(prds))
          names(nu) <- prds
          nu[names(tb)] <- tb
          nu
        }))
  df <- data.frame(users=users, df, stringsAsFactors=FALSE)
  df
}
#eg usage
prds <- c("morn.early", "morn", "aftn", "aftn.late", "evn", "nite", "nite.late")
updu <- user.prdusg(trips.sample, prds)
```

##Intermittency of use
```{r userinter}
intmtcOfUser <- function(user, trips){
  #print(paste("computing intermittency for user", user))
	trips.u <- subset(trips, user_id == user)
  trips.u <- trips.u[order(trips.u$start_time),]
  start.u <- trips.u$start_time
  end.u <- trips.u$end_time
  n <- nrow(trips.u)
  if(n > 1) list(user=user, intmtc =difftime(start.u[2:n], end.u[1:(n-1)], units="secs"))
  else c()
}
                                
user.intmtc <-  function(trips,
                         users=NULL,
                         statistics=TRUE,
                         .funcs=c(mean, var),
                         .funcs.names=c("mean", "var")){
  #.func.names needs some improvement, as now it wont generalize
  if(is.null(users)) users <- unique(trips$user_id)
  uintmt <- lapply(users, function(u) {
    intmtcOfUser(u, trips)$intmtc }
	)
  names(uintmt) <- users
  if (statistics){
    df <- as.data.frame(do.call(rbind, lapply(uintmt, function(u) {
        if(length(u) == 0) rep(NA, length(.funcs))
        else sapply(.funcs, function(f) f(u))
      }))
    )
    df <- data.frame(user_id=users, df)
    names(df)[2:3] <- .funcs.names
    df
  }
  else {
    do.call(rbind, lapply(1:length(users), function(i){
        data.frame(user=rep(users[i], length(uintmt[[i]])), intmtc=uintmt[[i]])
    }))
  }
}
  
uintmtc <- user.intmtc(trips=trips.sample)  
```  
##Duration of trips
Duration of trips can be easily extracted from *trips*
```{r userdurs}
user.durations <- function(trips,
                           statistics=TRUE,
                           .funcs=c(mean, var),
                           .funcs.names=c("mean", "var")){
  #.func.names needs some improvement, as now it wont generalize
	users <- unique(trips$user_id)
  udurs <- lapply(users, function(u) {
    trips.u <- subset(trips, user_id == u)
    difftime(trips.u$end_time, trips.u$start_time, units="secs")
  })
  if (statistics){
    df <- as.data.frame(do.call(rbind, lapply(udurs, function(u) {
        if(length(u) == 0) rep(NA, length(.funcs))
        else sapply(.funcs, function(f) f(u))
      }))
    )
    df <- data.frame(user_id=users, df)
    names(df)[2:3] <- .funcs.names
    df
  }
  else udurs
}
```
And now we can use the function to obtain the durations of user's trips,

```{r eguserdur}
udurs <- user.durations(trips=trips.sample)  
```  
##Diversity of stations
There aere three diversity values we can define for the stations that a user frequents, start, end and the start-to-end stations of each trip.

```{r userstartstatn}
#remove this line
trxns <- trxns.sample
entropy <- function(ps){
  ps <- ps/sum(ps)
  ps <- ps[ps > 0]
  -sum(ps*log2(ps))
}
user.stnentropies <- function(trxns){
  users <- unique(trxns$user_id)
  data.frame(user_id=users, 
             strtent = sapply(users, function(u){
																trxns.u <- subset(trxns, user_id == u)
																ss <- subset(trxns.u, drxn == "out")$statn_id
																entropy(table(ss))
															}),
             endent = sapply(users, function(u){
																trxns.u <- subset(trxns, user_id == u)
																es <- subset(trxns.u, drxn == "in")$statn_id
																entropy(table(es))
															}),
             strtend = sapply(users, function(u){
																trxns.u <- subset(trxns, user_id == u)
																ss <- subset(trxns.u, drxn == "out")$statn_id
																es <- subset(trxns.u, drxn == "in")$statn_id
																ses <- paste(ss, es, sep=".")
																entropy(table(ses))
															})
						 )
}
  
uents <- user.stnentropies(trxns=trxns)  
```


### Gathered user features
Ok, now lets put all that we have together,
```{r featureDataSet}
user.features <- merge(ufreqs, uintmtc, by="user_id")
user.features <- merge(user.features, udurs, by="user_id")
user.features <- merge(user.features, uents, by="user_id")
```
We can now cluster the users,
```{r usrclus}
user.ftr.sc <- scale(user.features[,-1])
user.ftr.sc <- data.frame(user_id=user.features$user_id, user.ftr.sc)
user.ftr.dis <- dist(user.ftr.sc[,-1], method="euclidean")
user.ftr.mat <- as.matrix(user.ftr.sc[,-1])
hc <- hclust(user.ftr.dis , method="ward.D2")
plot(hc)
#heatmap(user.ftr.mat)
user.ftr.sc <- user.ftr.sc[ order(cutree(hc, h=20)),]
ufm.m <- melt(user.ftr.sc[, -2], id.vars = "user_id")
p <- ggplot(ufm.m, aes(variable, user_id)) + 
      geom_tile(aes(fill=sign(value)*log10(abs(value))), color="white") +
       scale_fill_gradient(low="yellow", high="red")
print(p)
```
