---
title: "Variable naming scheme"
author: "Vishal Sood"
date: "13 Mar 2015"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
knitr::opts_chunk$set(echo=FALSE)
```

##Introduction
After looking at bike-sharing data from across the world we realise the features that we will require for our analysis. A given data set might call these features by a different set of names. We need a function to convert the names to the naming convention that we will use to right our analyses code.

```{r setvarnames}
namingScheme <- function(nms.orig,
			 var.user_id=NA,
			 var.trip_id=NA,
			 var.start_statn=NA,
			 var.end_statn=NA,
			 var.bike_id=NA,
			 var.start_time=NA,
			 var.end_time=NA){

	nms.rplc <- nms.orig
	names(nms.rplc) <- nms.orig

	nms.touse  <- c("user_id", "trip_id", "bike_id", 
			"start_statn", "end_statn",
			"start_time", "end_time")
			
	nms.torplc <- c(var.user_id, var.trip_id, var.bike_id,
			var.start_statn, var.end_statn,
			var.start_time, var.end_time)
	nms.touse <- nms.touse[ !is.na(nms.torplc)]
	nms.torplc <- nms.torplc[ !is.na(nms.torplc)]
	nms.rplc[ nms.torplc] <- nms.touse
	nms.rplc
}
```
We can test this function on the hubway data. To make this function code, we should write some tests !!!

	

                  
