---
title: 'Analysis Change of the Velopass data: the Bird''s Eye View'
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
  html_document:
    highlight: tango
    number_sections: yes
    theme: journal
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=FALSE)
knitr::opts_chunk$set(echo=FALSE)
knitr::opts_chunk$set(include=TRUE)
knitr::opts_chunk$set(results="asis")
knitr::opts_chunk$set(fig.width=12)
knitr::opts_chunk$set(fig.height=14)
prjpath <- "~/work/citiviz/bikeSharing/bikestats/"
datapath <- paste(prjpath, "data/velopass_data/", sep="")
analpath <- paste(prjpath, "analyses/", sep="")
rcodepath <- paste(analpath, "Rcode/", sep="")
setwd(analpath)
library(lattice)
```
```{r loadLibsAndSource, include=FALSE}
reqdpkgs <- c("lattice", "latticeExtra", "ggplot2", "reshape2",
						 "ggmap",	"plyr", "corrplot",
							"Hmisc", "corrplot", "rgdal", "sp")
lapply(reqdpkgs, library, character.only=TRUE)
source(paste(rcodepath, "datamanip.R", sep=""))
```




#Introduction
We look at the higher level properties of the Velopass data.

#Data
We have written the data into files on the disk, and read them from there,
```{r readdata}
dd <- paste(datapath, "transactions_lausanne-morges_2010/", sep="")
vlp2010 <- read.csv(paste(dd, "velopass_lausanne-morges_combined.csv", sep=""),
									 	stringsAsFactors=FALSE)
```
Once read, we have to convert the dates to date/time objects.
```{r datetime}
vlp2010 <- convertDateTime(vlp2010, c("start_time", "end_time"))
vlp2010 <- vlp2010[ order(vlp2010$start_time), ]
```

#Usage over time
How has the usage changed over time? Lets count the number of trips made every day.
```{r tripsbyday}
iswknd <- c(TRUE, rep(FALSE, 5), TRUE)
names(iswknd) <- c("Sunday", "Monday", "Tuesday", "Wednesday", 
											"Thursday", "Friday", "Saturday")
tripsByDay <- as.data.frame(table(floor_date(vlp2010$start_time, unit="day")),
													 	stringsAsFactors=FALSE)
names(tripsByDay) <- c("day", "num.trips")
tripsByDay$dayOfWeek <- factor(weekdays(ymd(tripsByDay$day)),
															 levels=c("Sunday", "Monday", "Tuesday",
																			 	"Wednesday", "Thursday", "Friday",
																			 	"Saturday") )
											
tripsByDay$day <- ymd(tripsByDay$day)
tripsByDay$weekdayOrEnd <- with(tripsByDay,
																		c("weekday", "weekend")[iswknd[dayOfWeek] + 1])
```
The data frame *tripsByDay* contains three columns,
```{r peektripsbyday, results="markup"}
print(head(tripsByDay))
```
We can plot the number of trips made on each day of the week, over the duration of the data set

```{r plottripsbyday}

xyplot( num.trips ~ day, group=dayOfWeek,
			  data=tripsByDay, 
        type="b",
				auto.key=TRUE,
        main="Number of trips each day")
```
Since people's activity patterns change between the weekdays and weekends, we can aggregate a week's days into these two categories,

```{r plottripsweekdayend}
xyplot( num.trips ~ day, group=weekdayOrEnd,
			  data=tripsByDay, 
        type="b",
				auto.key=TRUE,
        main="Number of trips each day")

```

We can add a column for the week during which the trip was made, and look at the number of trips by week of the year.

```{r tripsbyweek}

tripsByDay$week <- floor_date(ymd(tripsByDay$day), unit="week")
tripsByWeek <- ddply(tripsByDay, .(week, weekdayOrEnd), summarise, 
														 num.trips = sum(num.trips))
xyplot( num.trips ~ week, group=weekdayOrEnd,
			  data=tripsByWeek, 
        type="b",
				auto.key=TRUE,
        main="Number of trips each week")
```
And do the same by month.

```{r tripsbymonth}
tripsByDay$month <- floor_date(ymd(tripsByDay$day), unit="month")
tripsByMonth <- ddply(tripsByDay, .(month, weekdayOrEnd), summarise, 
														 num.trips = sum(num.trips))
xyplot( num.trips ~ month, group=weekdayOrEnd,
			  data=tripsByMonth, 
        type="b",
				auto.key=TRUE,
        main="Number of trips each month")
```

We see significant variation in usage over the months, but there may be a pattern of weekly usage independent of the month. One pattern is that the usage on weekdays is much larger than that on weekends. The usage for individual days is hidden under noise, which we can handle by summing up all days in a month.

```{r tripsbydayaggregatemonth}
tripsByWeekday.month <- ddply(tripsByDay, .(month, dayOfWeek), summarise,
													 mean.num.trips = mean(num.trips),
													 var.num.trips = var(num.trips) )

tripsByWeekday.month$lower.num.trips <- with(tripsByWeekday.month,
																						 mean.num.trips - sqrt(var.num.trips))
tripsByWeekday.month$upper.num.trips <- with(tripsByWeekday.month,
																						 mean.num.trips + sqrt(var.num.trips))

tripsByWeekday.month$num.dayOfWeek <- as.numeric(tripsByWeekday.month$dayOfWeek)
xYplot(Cbind(mean.num.trips, lower.num.trips, upper.num.trips) ~ num.dayOfWeek | month,
			 data=tripsByWeekday.month,
			 type='b',
			 main="Number of trips for day of the week")
```

We have assumed a normal distribution for the number of trips during a day to plot the error bars as the standard deviation from the mean. However we should expect number of trips in a day to be Poisson distributed.

```{r histtripsdistbn}
histogram(~ num.trips | weekdayOrEnd,
					data=tripsByDay,
					main="Histograms of number of trips in a day")

histogram(~ num.trips | month,
					data=subset(tripsByDay, weekdayOrEnd == "weekday"),
					main="Histograms of number of trips in a day for weekdays")

histogram(~ num.trips | month,
					data=subset(tripsByDay, weekdayOrEnd == "weekend"),
					main="Histograms of number of trips in a day for weekends")
```




We can look for number of trips during certain hours of the day,  summed/averaged over days/weeks/months/years. We first need a function that converts time of day to seconds.

```{r timeofdayinsecs}
timeOfDayInSeconds <- function(ts, format){
	if (all(class(ts) == c("POSIXct","POSIXt"))){
		3600*hour(ts) + 60*minute(ts) + second(ts)
	}
	else if (class(ts) == "character"){
		if (format == "hms"){
			sapply(ts, function(t) {
							sum(c(3600, 60, 1) * as.numeric(strsplit(x=t, split=":")[[1]]))
						})
		}
		else if (format == "ymd hms"){
			ts <- do.call(rbind, strsplit(x=ts, split=" "))[,2]
			timeOfDayInSeconds( ts, format="hms")
		}
		else {
			stop("UNKNOWN FORMAT: allowed 'hms', 'ymd hms'")
		}
	}
	else {
		stop("t of UNKNOWN CLASS: allowed character and POSIXct/t")
	}
}
```


```{r tripsAtHour}

tripsStartingInPeriod <- function(trips, start_prd, end_prd, format="hms"){
	spsecs = timeOfDayInSeconds(start_prd, format="hms")
	epsecs = timeOfDayInSeconds(end_prd, format="hms")
	stsecs = timeOfDayInSeconds(trips$start_time)
	trips[ stsecs >= spsecs & stsecs <= epsecs,]
}

tripsEndingInPeriod <- function(trips, start_prd, end_prd, format="hms"){
	spsecs = timeOfDayInSeconds(start_prd, format="hms")
	epsecs = timeOfDayInSeconds(end_prd, format="hms")
	etsecs = timeOfDayInSeconds(trips$end_time)
	trips[ etsecs >= spsecs & etsecs <= epsecs,]
}

withByWeekAndMonth <- function(tbd, varDateTime="day"){
	cls <- class(tbd[, varDateTime])
	if (all(cls == c("POSIXct", "POSIXt"))) {
		tbd$day = floor_date( tbd[ , varDateTime], unit="day")
	}
	else if (cls != "character") {
		stop(paste("class", class(tbd), "of provided data not implemented"))
	}
	tbd$day <- ymd(tbd$day)
	tbd$dayOfWeek <- factor(weekdays(tbd[, varDateTime]), 
													levels=c("Sunday", "Monday", "Tuesday", 
																	 "Wednesday", "Thursday", "Friday", "Saturday") )
	tbd$wkdayOrEnd <- c("weekday", "weekend")[iswknd[tbd$dayOfWeek] + 1]
	tbd$week <- floor_date(tbd[, varDateTime], unit="week")
	tbd$month <- floor_date(tbd[, varDateTime], unit="month")
	tbd
}


countTripsBwnHours <- function(trips, start_prd, end_prd, by="day", plot=TRUE){
                    
	starting <- tripsStartingInPeriod(trips, start_prd, end_prd)
	ending <- tripsEndingInPeriod(trips, start_prd, end_prd)
	startingByDay <- as.data.frame(table(floor_date(starting$start_time, unit="day")),
																 stringsAsFactors=FALSE)
	names(startingByDay) <- c("day", "num.trips")
	startingByDay <- withByWeekAndMonth(startingByDay)
	endingByDay <- as.data.frame(table(floor_date(ending$start_time, unit="day")),
																 stringsAsFactors=FALSE)
	names(endingByDay) <- c("day", "num.trips")
	endingByDay <- withByWeekAndMonth(endingByDay)
	counts <- rbind(data.frame(ddply(startingByDay, c(by, "wkdayOrEnd"), summarise,
																		num.trips = sum(num.trips)),
															drxn="out", start_prd=start_prd, end_prd=end_prd),
									data.frame(ddply(endingByDay, c(by, "wkdayOrEnd"), summarise,
																	 num.trips = sum(num.trips)),
														 drxn="in", start_prd=start_prd, end_prd=end_prd) 
									)

	counts.p <- counts
	counts.p$period <- counts[, by]
	if(plot){
		p  <-  xyplot(num.trips ~ period | drxn, group=wkdayOrEnd, 
								 data=counts.p,
								 type="b",
								 auto.key=TRUE,
								 main=paste("total trips by ", by, " between ", 
														start_prd, " and ", end_prd, sep="") 
								)
		print(p)
	}
	counts
}

```
We can use the functions to count trips to find trips by the hour. First we write a function that will get the counts for every hour of the day, and include an offset in minutes.

```{r cth}
countTripsHourly <- function(trips, offset = 0, by="day", plot=FALSE) {
	hrs <- 	0:23 
	do.call(rbind, 
					lapply(0:23, function(h) countTripsBwnHours(trips, paste(h, offset, 0, sep=":"),
																										  paste(h+1, offset, 0, sep=":"),
																										  by=by, plot=plot))
				 )
}
```

Now we can make some plots for usage by the hour over the entire time duration of the data.

```{r dailyusagebythehour}
usageByTheHour.aggDay <- countTripsHourly(vlp2010)

xyplot( num.trips ~ day | start_prd,
			 	group=drxn,
			  data=subset(usageByTheHour.aggDay, wkdayOrEnd == "weekend"),
				auto.key=TRUE,
				type="b",
				main = "Usage by the hour on weekends")

xyplot( num.trips ~ day | start_prd,
			 	group=drxn,
			  data=subset(usageByTheHour.aggDay, wkdayOrEnd == "weekday"),
				auto.key=TRUE,
				type="b",
				main = "Usage by the hour on weekdays")

```

So far we have looked at the behavior of number of trips in a day. What we would also want to see how many trips are made over a certain period of the day. We can use the data produced by *countTripsHourly* to make these plots as well,

```{r hourlyusage}
hourlyUsage <- withByWeekAndMonth(vlp2010, "start_time")
hourlyUsage$start_prd <- hour(hourlyUsage$start_time)
hourlyUsage$end_prd <- hour(hourlyUsage$end_time)
hourlyOutUsage.aggYear <- data.frame(ddply(hourlyUsage, .(start_prd), .fun="nrow"), 
															 					 drxn="out")
names(hourlyOutUsage.aggYear) <- c("hour", "num.trips", "drxn")
hourlyInUsage.aggYear <- data.frame(ddply(hourlyUsage, .(end_prd), .fun="nrow"),
																		drxn="in")
names(hourlyInUsage.aggYear) <- c("hour", "num.trips", "drxn")

hourlyUsage.aggYear <- rbind(hourlyOutUsage.aggYear, hourlyInUsage.aggYear)

xyplot( num.trips ~ hour| drxn,
			 data=hourlyUsage.aggYear,
			 type="b")
```

Above we saw total hourly usage for the whole year (all the data in *vlp2010*), and we can aggregate the trips for each month, or for each week as well.

```{r hourlyusagemonth}
hourlyOutUsage.aggMonth <- data.frame(ddply(hourlyUsage, .(start_prd, wkdayOrEnd, month), .fun="nrow"),
																						drxn="out")
names(hourlyOutUsage.aggMonth) <- c("hour", "wkdayOrEnd", "month", "num.trips", "drxn")
hourlyInUsage.aggMonth <- data.frame(ddply(hourlyUsage, .(end_prd, wkdayOrEnd, month), .fun="nrow"),
																						drxn="in")
names(hourlyInUsage.aggMonth) <- c("hour", "wkdayOrEnd", "month", "num.trips", "drxn")
hourlyUsage.aggMonth <- rbind(hourlyOutUsage.aggMonth, hourlyInUsage.aggMonth)
hourlyUsage.aggMonth <- subset(hourlyUsage.aggMonth, !is.na(hour))
xyplot(num.trips ~ hour | month, 
			 group=drxn,
			 data=subset(hourlyUsage.aggMonth, wkdayOrEnd=="weekday"),
			 auto.key=TRUE, type="b",
			 main="Hourly usage on weekdays over a month")

xyplot(num.trips ~ hour | month, 
			 group=drxn,
			 data=subset(hourlyUsage.aggMonth, wkdayOrEnd=="weekend"),
			 auto.key=TRUE, type="b",
			 main="Hourly usage on weekends over a month")
```

###Hourly usage for a particular station

We can produce the hourly usage plots for a particular station,
```{r hourlystationusageAggYear}
		stoe <- subset(vlp2010,
									 start_statn == "0101 Place de l'Europe")
		hourlyStoe <- withByWeekAndMonth(stoe, "start_time")
		hourlyStoe$start_prd <- hour( hourlyStoe$start_time)
		hourlyStoe$end_prd <- hour( hourlyStoe$end_time)
		hourlyStoeOut.aggYear <- data.frame(ddply(hourlyStoe, .(start_prd, wkdayOrEnd), .fun="nrow"),
																			 drxn="out")
		names(hourlyStoeOut.aggYear) <- c("hour", "wkdayOrEnd", "num.trips", "drxn")
		hourlyStoeIn.aggYear <- data.frame(ddply(hourlyStoe, .(end_prd, wkdayOrEnd), .fun="nrow"),
																			  drxn="in")
		names(hourlyStoeIn.aggYear) <- c("hour", "wkdayOrEnd", "num.trips", "drxn")
		hourlyStoe.aggYear <- rbind(hourlyStoeOut.aggYear, hourlyStoeIn.aggYear)
	
		xyplot(num.trips ~ hour, 
					 group=drxn,
					 data=subset(hourlyStoe.aggYear, wkdayOrEnd=="weekday"),
					 auto.key=TRUE,
					 type="b",
					 main="Hourly weekday usage, aggregated over a year")
```

And aggregated over a month,

```{r hourlystationusageAggMonth}

		## ----hourlyusagemonth----------------------------------------------------
		hourlyStoeOut.aggMonth <- data.frame(ddply(hourlyStoe, 
																							 .(start_prd, wkdayOrEnd, month), .fun="nrow"),
																				 drxn="out")
		names(hourlyStoeOut.aggMonth) <- c("hour", "wkdayOrEnd", "month", "num.trips", "drxn")
		hourlyStoeIn.aggMonth <- data.frame(ddply(hourlyStoe, 
																							.(end_prd, wkdayOrEnd, month), .fun="nrow"),
																				drxn="in")
		names(hourlyStoeIn.aggMonth) <- c("hour", "wkdayOrEnd", "month", "num.trips", "drxn")
		hourlyStoe.aggMonth <- rbind(hourlyStoeOut.aggMonth, hourlyStoeIn.aggMonth)
		hourlyStoe.aggMonth <- subset(hourlyStoe.aggMonth, !is.na(hour))

		xyplot(num.trips ~ hour | month,
					 group=drxn,
					 data=subset(hourlyStoe.aggMonth, wkdayOrEnd=="weekday"),
					 auto.key=TRUE,
					 type="b",
					 main="Hourly weekday usage, aggregated over a month")
```


##Trip durations

```{r tripdurations}
vlp2010 <- withByWeekAndMonth(vlp2010, "start_time")
vlp2010$duration <- duration(new_interval(vlp2010$start_time, vlp2010$end_time), units="seconds")
histogram( ~ log10(duration) | wkdayOrEnd,
				 	 data=vlp2010,
					 type="count",
					 main="Histogram for the duration of trips")

histogram( ~ log10(duration) | month,
           data=subset(vlp2010, wkdayOrEnd=="weekday"),
					 type="count",
					 main="Histogram for the duration of trips on weekdays")

histogram( ~ log10(duration) | month,
           data=subset(vlp2010, wkdayOrEnd=="weekend"),
					 type="count",
					 main="Histogram for the duration of trips on weekends")
```

##Statistics for stations
What stations are used most often?
```{r stationUsage}
histogram( ~ factor(start_statn) | wkdayOrEnd,
					data=vlp2010, 
					type="count",
					auto.key=TRUE,
					scales=list(x=list(rot=90)),
					main="Histogram for station usage for outgoing trips")

histogram( ~ factor(end_statn) | wkdayOrEnd,
					data=vlp2010, 
					type="count",
					auto.key=TRUE,
					scales=list(x=list(rot=90)),
					main="Histogram for station usage for incoming trips")
```

Individual station usage statistics is easy. We should also count the number of stations between each pair of stations. For this we can make a table.

```{r trafficbwnstns}
traffic <- table(vlp2010$start_statn, vlp2010$end_statn)
```

The best way to explore the traffic between stations is to map them. For this we need the geolocations of the stations.

```{r stnloc}
stnlocs <- read.csv(paste(datapath, "stations.csv", sep=""), sep=",", 
										stringsAsFactors=FALSE)
stnlocs.spatial <- stnlocs
coordinates(stnlocs.spatial) <- c("lon", "lat")
proj4string(stnlocs.spatial) <- CRS("+init=epsg:4326")
```
We use google to provide an underlay
```{r stnmap}
b <- bbox(stnlocs.spatial)
b[1, ] <- (b[1,] - mean(b[1,])) * 1.05 + mean(b[1,])
b[2, ] <- (b[2,] - mean(b[2,])) * 1.05 + mean(b[2,])
stn.map <- ggmap(get_map(location=b, source="stamen", maptype="toner", crop=TRUE))
stn.map <- ggmap(get_map(location=b, source="google",  crop=TRUE, zoom=14))
stn.map + geom_point(data=stnlocs, aes(x=lon, y=lat), size=6, col="red")

```

