---
title: 'Preparation of the Velopass data"
output:
  pdf_document: default
  html_document:
    highlight: tango
    number_sections: yes
    theme: journal
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=FALSE)
knitr::opts_chunk$set(echo=FALSE)
knitr::opts_chunk$set(include=TRUE)
knitr::opts_chunk$set(results="asis")
knitr::opts_chunk$set(fig.width=12)
knitr::opts_chunk$set(fig.height=14)
prjpath <- "~/work/citiviz/bikeSharing/bikestats/"
datapath <- paste(prjpath, "data/velopass_data/", sep="")
analpath <- paste(prjpath, "analyses/", sep="")
rcodepath <- paste(analpath, "Rcode/", sep="")
setwd(analpath)
library(lattice)
```
```{r loadLibsAndSource, include=FALSE}
reqdpkgs <- c("lattice", "latticeExtra", "ggplot2", "reshape2", "plyr", "corrplot",
							"Hmisc", "corrplot")
lapply(reqdpkgs, library, character.only=TRUE)
source(paste(rcodepath, "datamanip.R", sep=""))
```

#Introduction
We prepare the velopass data. There are three sub-folders containing data for Velopass, each containing more than one csv. We load the csvs and combine them into one data csvs for each sub-folder

#Data collation

##Transactions campus 
There are 6 pages of data in *transactions_campus_2010* for the year 2010. We will create a single data-frame from them.
```{r datacollation}
dd <- paste(datapath, "transactions_campus_2010/", sep="")
vlp2010.1 <- read.csv(paste(dd, "transactions_page1_2010.csv", sep=""), 
											sep=";",	stringsAsFactors=FALSE) 
vlp2010.2 <- read.csv(paste(dd, "transactions_page2_2010.csv", sep=""), 
											sep=";",	stringsAsFactors=FALSE) 
vlp2010.3 <- read.csv(paste(dd, "transactions_page3_2010.csv", sep=""), 
											sep=";",	stringsAsFactors=FALSE) 
vlp2010.4 <- read.csv(paste(dd, "transactions_page4_2010.csv", sep=""), 
											sep=";",	stringsAsFactors=FALSE) 
vlp2010.5 <- read.csv(paste(dd, "transactions_page5_2010.csv", sep=""), 
											sep=";",	stringsAsFactors=FALSE) 
vlp2010.6 <- read.csv(paste(dd, "transactions_page6_2010.csv", sep=""), 
											sep=";",	stringsAsFactors=FALSE) 
```

There is a column "X" in some of the data sets, which we will not include in our unified data frame. 
```{r datacollation2}
vlp2010.list <- list(vlp2010.1, vlp2010.2, vlp2010.3, vlp2010.4, vlp2010.5, vlp2010.6)
vlp2010 <- do.call(rbind, lapply(vlp2010.list, function(vlpy) vlpy[, -6]))
```
We also need to use our naming scheme for the variables.
```{r uniformNamingScheme}
source(paste(rcodepath, "varNamingScheme.R", sep=""))
nms.indata <- names(vlp2010)
names(vlp2010) <- namingScheme(nms.indata,
															 var.user_id="user_id",
															 var.start_statn="from_station",
															 var.end_statn="to_station",
															 var.start_time="from_timestamp",
															 var.end_time="to_timestamp",
															 var.trip_id=NA)

```



The dates in _vlp2010_ are if the form "MON dd - h:m:s". To use our libraries we will have translate the Italian month names in MON to English, and prepend the dates by year.

```{r italianMonthToEnglish}
vlpDates <- rbind(vlp2010$start_time, vlp2010$end_time)
vlpMonths <- unique( sapply(strsplit(x=vlpDates, split=" "), function(v) v[2]))

itlToEngMonths <- data.frame(italian=c("Gen", "Feb", "Mar", "Apr", "Mag", "Giu",
																			 "Lug", "Ago", "Set", "Ott", "Nov", "Dic"),
															english=c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
																				"Jul", "Aug", "Sep", "Oct", "Nov", "Dec"),
														 index=1:12)
itlToEngMonths$italian <- as.character(itlToEngMonths$italian)
itlToEngMonths$english <- as.character(itlToEngMonths$english)
rownames(itlToEngMonths) <- itlToEngMonths$italian


convertVLPdate <- function(date){
	dsplit <- do.call(rbind, strsplit(x=date, split=" "))
	ymd <- paste(2010,  itlToEngMonths[dsplit[,2], "index"], dsplit[,1], sep="-")
	paste(ymd, dsplit[,5], sep=" ")
}

vlp2010$start_time <- convertVLPdate(vlp2010$start_time)
vlp2010$end_time <- convertVLPdate(vlp2010$end_time)

vlp2010 <- convertDateTime(vlp2010, c("start_time", "end_time"))

```

Some of the code below uses gender, which is not available in the data,
```{r addgender}
vlp2010$gender <- "Both"
```

Lets save the combined data-frame to a file,
```{r writeCombinedDataSet}
vlp2010.towrite <- vlp2010
vlp2010.towrite$start_time <- as.character(vlp2010.towrite$start_time)
vlp2010.towrite$end_time <- as.character(vlp2010.towrite$end_time)

write.csv(vlp2010, paste(dd, "velopass_combined.csv", sep=""), row.names=FALSE)

```

The data in _vlp2010_ is for several months. Let us begin our analysis by looking at a single month's worth of data,
```{r sampleMonth}
source(paste(rcodepath, "datamanip.R",sep=""))
data.sample <- extractGivenDays(vlp2010,
																start=ymd_hms("2010-06-01 00:00:00"),
																end=ymd_hms("2010-06-15 00:00:00"))

#data.sample <- vlp2010[ months(vlp2010$start_time) == "June", ]

```

##Transactions Lausanne-Morges 2010
There are two csvs.

```{r dataload}
dd <- paste(datapath, "transactions_lausanne-morges_2010", sep="/")
vlplm2010 <- rbind(read.csv(paste(dd, "transactions_page1_2010.csv", sep="/"), sep=";", 
													  stringsAsFactors=FALSE	),
									 read.csv(paste(dd, "transactions_page2_2010.csv", sep="/"), sep=";", 
										    		stringsAsFactors=FALSE	)
									)
```

Having loaded the data, we will migrate the variable names to use our uniform naming convention,
```{r unifynameslm}
source(paste(rcodepath, "varNamingScheme.R", sep=""))
nms.indata <- names(vlplm2010)
names(vlplm2010) <- namingScheme(nms.indata,
																 var.user_id="user_id",
																 var.start_statn="from_station",
																 var.end_statn="to_station",
																 var.start_time="from_timestamp",
																 var.end_time="to_timestamp",
																 var.trip_id=NA)
```
The dates in _vlp2010_ are if the form "MON dd - h:m:s". To use our libraries we will have to translate the Italian month names in MON to English, and prepend the dates by year.

```{r italianMonthToEnglishlm}
vlplm2010$start_time <- convertVLPdate(vlplm2010$start_time)
vlplm2010$end_time <- convertVLPdate(vlplm2010$end_time)
vlplm2010 <- convertDateTime(vlplm2010, c("start_time", "end_time"))
```

We will now save the data in *vlplm2010* to a file,
```{r writevlplm2010}
vlp2010.towrite <- vlplm2010
vlp2010.towrite$start_time <- as.character(vlp2010.towrite$start_time)
vlp2010.towrite$end_time <- as.character(vlp2010.towrite$end_time)

write.csv(vlp2010.towrite, 
					paste(dd, "velopass_lausanne-morges_combined.csv", sep="/"), 
					row.names=FALSE)
```



