---
title: "Analysis of the Velopass data"
output:
  html_document:
    highlight: tango
    number_sections: yes
    theme: journal
  pdf_document: default
---
```{r setup, include=TRUE}
knitr::opts_chunk$set(cache=FALSE)
knitr::opts_chunk$set(echo=TRUE)
knitr::opts_chunk$set(fig.width=14)
knitr::opts_chunk$set(fig.height=14)
prjpath <- "~/work/citiviz/bikeSharing/bikestats/"
datapath <- paste(prjpath, "data/velopass_data/", sep="")
analpath <- paste(prjpath, "analyses/", sep="")
rcodepath <- paste(analpath, "Rcode/", sep="")
setwd(analpath)
library(lattice)
```


#Introduction
We analyze the data from velopass.

#Data collation
There are 6 pages of data for the year 2010. We will create a single data-frame from them.
```{r datacollation}
dd <- paste(datapath, "transactions_campus_2010/", sep="")
vlp2010.1 <- read.csv(paste(dd, "transactions_page1_2010.csv", sep=""), 
											sep=";",	stringsAsFactors=FALSE) 
vlp2010.2 <- read.csv(paste(dd, "transactions_page2_2010.csv", sep=""), 
											sep=";",	stringsAsFactors=FALSE) 
vlp2010.3 <- read.csv(paste(dd, "transactions_page3_2010.csv", sep=""), 
											sep=";",	stringsAsFactors=FALSE) 
vlp2010.4 <- read.csv(paste(dd, "transactions_page4_2010.csv", sep=""), 
											sep=";",	stringsAsFactors=FALSE) 
vlp2010.5 <- read.csv(paste(dd, "transactions_page5_2010.csv", sep=""), 
											sep=";",	stringsAsFactors=FALSE) 
vlp2010.6 <- read.csv(paste(dd, "transactions_page6_2010.csv", sep=""), 
											sep=";",	stringsAsFactors=FALSE) 
```

There is a column "X" in some of the data sets, which we will not include in our unified data frame. 
```{r datacollation2}
vlp2010.list <- list(vlp2010.1, vlp2010.2, vlp2010.3, vlp2010.4, vlp2010.5, vlp2010.6)
vlp2010 <- do.call(rbind, lapply(vlp2010.list, function(vlpy) vlpy[, -6]))
```
We also need to use our naming scheme for the variables.
```{r uniformNamingScheme}
source(paste(rcodepath, "varNamingScheme.R", sep=""))
nms.indata <- names(vlp2010)
names(vlp2010) <- namingScheme(nms.indata,
															 var.user_id="user_id",
															 var.start_statn="from_station",
															 var.end_statn="to_station",
															 var.start_time="from_timestamp",
															 var.end_time="to_timestamp",
															 var.trip_id=NA)

```



The dates in _vlp2010_ are if the form "MON dd - h:m:s". To use our libraries we will have translate the Italian month names in MON to English, and prepend the dates by year.

```{r italianMonthToEnglish}
vlpDates <- rbind(vlp2010$start_time, vlp2010$end_time)
vlpMonths <- unique( sapply(strsplit(x=vlpDates, split=" "), function(v) v[2]))

itlToEngMonths <- data.frame(italian=c("Gen", "Feb", "Mar", "Apr", "Mag", "Giu",
																			 "Lug", "Ago", "Set", "Ott", "Nov", "Dec"),
															english=c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
																				"Jul", "Aug", "Sep", "Oct", "Nov", "Dec"),
														 index=1:12)
itlToEngMonths$italian <- as.character(itlToEngMonths$italian)
itlToEngMonths$english <- as.character(itlToEngMonths$english)
rownames(itlToEngMonths) <- itlToEngMonths$italian


convertVLPdate <- function(date){
	dsplit <- strsplit(x=date, split=" ")[[1]]
	ymd <- paste(2010,  itlToEngMonths[dsplit[2], "index"], dsplit[1], sep="-")
	paste(ymd, dsplit[5], sep=" ")
}

vlp2010$start_time <- sapply(vlp2010$start_time, convertVLPdate)
vlp2010$end_time <- sapply(vlp2010$end_time, convertVLPdate)

vlp2010 <- convertDateTime(vlp2010, c("start_time", "end_time"))

```

Some of the code below uses gender, which is not available in the data,
```{r addgender}
vlp2010$gender <- "Both"
```

Lets save the combined data-frame to a file,
```{r writeCombinedDataSet}
vlp2010.towrite <- vlp2010
vlp2010.towrite$start_time <- as.character(vlp2010.towrite$start_time)
vlp2010.towrite$end_time <- as.character(vlp2010.towrite$end_time)

write.csv(vlp2010, paste(dd, "velopass_combined.csv", sep=""), row.names=FALSE)

```

The data in _vlp2010_ is for several months. Let us begin our analysis by looking at a single month's worth of data,
```{r sampleMonth}
source(paste(rcodepath, "datamanip.R",sep=""))
data.sample <- extractGivenDays(vlp2010,
																start=ymd_hms("2010-06-01 00:00:00"),
																end=ymd_hms("2010-06-15 00:00:00"))

#data.sample <- vlp2010[ months(vlp2010$start_time) == "June", ]

```


#Analyses
The following is a paste from _userPerspective.Rmd_ that was written while analyzing hubway data. 

# Adding features to the data

Biker behavior in the morning can be different from that in the afternoon. We will divide each day into periods,
```{r dayperiods}
periodOfDay <- function(times){
  dayPeriods <- data.frame(
    period=c("morn.early", "morn", "aftn", "aftn.late", "evn", "nite", "nite.late"),
    start=c(5, 8, 12, 15, 18, 21, 0),
    end=c(8,12,15,18,21,24,5)
  )
  dayPeriods$period <- as.character(dayPeriods$period)
  if (class(times) != "integer") times <- hour(times)
  sapply(times, function(t) {
    with(dayPeriods, period[start <= t & end > t])
  })
}

data.sample$period <- periodOfDay(data.sample$start_time)
```  

#Transactions
We have written a function to convert trips to transactions (see _datamanip.Rmd_), that we use to convert the trips to transactions in the hubway sample.

The columns in the transactions are,
```{r transclmns}
names(makeTrxn(data.sample[1,]))
```
The column *statn* should provide the location of the bike-station, either directly as (long, lat) or as a label along with a dictionary that can be queried with that label.
From now on we will have two forms for the data, as trips, and as transactions.
```{r hwmaketrans}
trips.sample <- data.sample
trxns.sample <- makeTrxn(trips.sample)
trxns.sample$period <- periodOfDay( trxns.sample$time)
```

We can take a peek at the transactions in our sample week,
```{r transpeek}
head(trxns.sample)
```

Transalting transactions to trips is harder, and will be implemented later.

#Statistics for a single user
How often does a user use bikes? At what time of day? Which stations are most frequented by a user? What is the duration of a user's trips? 

```{r usagetrace}
usagetrace <- function(user, trips, dt=minutes(5)){
	userTrips <- subset(trips, user_id == user)
  #make a sequence by 5 minutes
  t0 <- min(userTrips$start_time)
  tl <- max(userTrips$end_time)
  N <- duration(interval(start=t0, end=tl))/as.duration(dt)
  tvec <- t0 + (0:N)*dt 
  tripIndices <- lapply(1:nrow(trips), function(i) {
    which( tvec > userTrips$start_time[i] & tvec < userTrips$end_time[i])
  })
  tripIndices <- do.call(c, tripIndices)
  inuse <- rep(0, length(tvec))
  inuse[tripIndices] <- 1
  data.frame(time=tvec, inuse=inuse, user_id=user)
}
```
Lets explore usage traces of some users,
```{r explusgtrcs, cache=TRUE}
allusers <- unique(trips.sample$user_id)
eg.ustrs <- do.call(rbind, lapply(allusers[1:64], function(u) usagetrace(u, trips.sample)))
xyplot( inuse ~ time | factor(user_id), data=eg.ustrs, type="l")
```

What about the intermittency of use by an individual user? 

#User features

In order to study the users, we extract features from their usage data. The features we will discuss are as follows,
1. Frequency of use
2. Duration of trips
3. Intermittency of use
4. The diversity of start stations
5. The diversity of end stations
6. The diversity of (start, end) station pairs.

To illustrate the features we will use both the trip and transaction forms of the data with disambiguated user ids, as our starting data. 
### Frequency of use
A user's frequency is the number of times they have used the bike-sharing system, or the number of trips they have made. We will assume that a user who rents a bike also returns it in the dataset.
```{r userfreq}
library(reshape2)
library(plyr)
user.frequencies <- function(trxns){
  trxn.types <- unique(trxns$drxn )
  users <- unique(trxns$user_id)
  tb <- as.data.frame(table(trxns$user_id, trxns$drxn))
  names(tb) <- c("user_id", "drxn", "freq")
  dcast(data=tb, formula=user_id ~ drxn, value.var="freq", fun.aggregate=sum)
}

ufreqs <- user.frequencies(trxns=trxns.sample)
```

Lets explore user frequencies,
```{r}
ufreqs.l <- melt(ufreqs, id.vars = "user_id", 
                 variable.name = "drxn", value.name="count" )
histogram(~ count | drxn, data=ufreqs.l)
```

## Frequency of use by time of day
We have the variable _period_ that we can use to study the period of the day when the bikes are used. 
```{r periodbars}
library(ggplot2)
print(names(trxns.sample))
prdUsg <- with(trxns.sample, as.data.frame(table(gender, period, drxn)))
levels(prdUsg$period) <- c("morn.early", "morn", 
                          "aftn", "aftn.late", 
                          "evn", "evn.late", 
                          "nite", "nite.late")

gp <- ggplot(data=prdUsg, aes(x=period, y=Freq, fill=gender)) +
        geom_bar(stat="identity", position="dodge") + 
          facet_grid( drxn ~ . ) + 
            ggtitle("Usage going out by period of day across genders")
print(gp)

gp.out <- ggplot(data=subset(prdUsg, drxn =="out"),
                             aes(x=period, y=Freq, fill=gender)) +
            geom_bar(stat="identity", position="dodge") + 
              ggtitle("Usage going out by period of day across genders")
print(gp.out)

gp.in <- ggplot(data=subset(prdUsg, drxn =="in"),
                             aes(x=period, y=Freq, fill=gender)) +
            geom_bar(stat="identity", position="dodge") +
              ggtitle("Usage coming in by period of day across genders")
print(gp.in)
```
We see that usage by both males and females follows the same pattern during the day. We can compute the frequency of use across the day for each user,
```{r userdayperiod}
user.prdusg <- function(data, 
                        prds){
  users <- unique(data$user_id)
  df <- do.call(rbind, lapply(users, function(u){
					tb <- table(subset(data, user_id == u)$period)
          nu <- rep(0, length(prds))
          names(nu) <- prds
          nu[names(tb)] <- tb
          nu
        }))
  df <- data.frame(users=users, df, stringsAsFactors=FALSE)
  df
}
#eg usage
prds <- c("morn.early", "morn", "aftn", "aftn.late", "evn", "nite", "nite.late")
updu <- user.prdusg(trips.sample, prds)
```

##Intermittency of use
```{r userinter}
intmtcOfUser <- function(user, trips){
  #print(paste("computing intermittency for user", user))
	trips.u <- subset(trips, user_id == user)
  trips.u <- trips.u[order(trips.u$start_time),]
  start.u <- trips.u$start_time
  end.u <- trips.u$end_time
  n <- nrow(trips.u)
  if(n > 1) list(user=user, intmtc =difftime(start.u[2:n], end.u[1:(n-1)], units="secs"))
  else c()
}
                                
user.intmtc <-  function(trips,
                         users=NULL,
                         statistics=TRUE,
                         .funcs=c(mean, var),
                         .funcs.names=c("mean", "var")){
  #.func.names needs some improvement, as now it wont generalize
  if(is.null(users)) users <- unique(trips$user_id)
  uintmt <- lapply(users, function(u) {
    intmtcOfUser(u, trips)$intmtc }
	)
  names(uintmt) <- users
  if (statistics){
    df <- as.data.frame(do.call(rbind, lapply(uintmt, function(u) {
        if(length(u) == 0) rep(NA, length(.funcs))
        else sapply(.funcs, function(f) f(u))
      }))
    )
    df <- data.frame(user_id=users, df)
    names(df)[2:3] <- .funcs.names
    df
  }
  else {
    do.call(rbind, lapply(1:length(users), function(i){
        data.frame(user=rep(users[i], length(uintmt[[i]])), intmtc=uintmt[[i]])
    }))
  }
}
  
uintmtc <- user.intmtc(trips=trips.sample)  
```  
##Duration of trips
Duration of trips can be easily extracted from *trips*
```{r userdurs}
user.durations <- function(trips,
                           statistics=TRUE,
                           .funcs=c(mean, var),
                           .funcs.names=c("mean", "var")){
  #.func.names needs some improvement, as now it wont generalize
	users <- unique(trips$user_id)
  udurs <- lapply(users, function(u) {
    trips.u <- subset(trips, user_id == u)
    difftime(trips.u$end_time, trips.u$start_time, units="secs")
  })
  if (statistics){
    df <- as.data.frame(do.call(rbind, lapply(udurs, function(u) {
        if(length(u) == 0) rep(NA, length(.funcs))
        else sapply(.funcs, function(f) f(u))
      }))
    )
    df <- data.frame(user_id=users, df)
    names(df)[2:3] <- .funcs.names
    df
  }
  else udurs
}
```
And now we can use the function to obtain the durations of user's trips,

```{r eguserdur}
udurs <- user.durations(trips=trips.sample)  
```  
##Diversity of stations
There aere three diversity values we can define for the stations that a user frequents, start, end and the start-to-end stations of each trip.

```{r userstartstatn}
#remove this line
trxns <- trxns.sample
entropy <- function(ps){
  ps <- ps/sum(ps)
  ps <- ps[ps > 0]
  -sum(ps*log2(ps))
}
user.stnentropies <- function(trxns){
  users <- unique(trxns$user_id)
  data.frame(user_id=users, 
             strtent = sapply(users, function(u){
																trxns.u <- subset(trxns, user_id == u)
																ss <- subset(trxns.u, drxn == "out")$statn_id
																entropy(table(ss))
															}),
             endent = sapply(users, function(u){
																trxns.u <- subset(trxns, user_id == u)
																es <- subset(trxns.u, drxn == "in")$statn_id
																entropy(table(es))
															}),
             strtend = sapply(users, function(u){
																trxns.u <- subset(trxns, user_id == u)
																ss <- subset(trxns.u, drxn == "out")$statn_id
																es <- subset(trxns.u, drxn == "in")$statn_id
																ses <- paste(ss, es, sep=".")
																entropy(table(ses))
															})
						 )
}
  
uents <- user.stnentropies(trxns=trxns)  
```


### Gathered user features
Ok, now lets put all that we have together,
```{r featureDataSet}
user.features <- merge(ufreqs, uintmtc, by="user_id")
user.features <- merge(user.features, udurs, by="user_id")
user.features <- merge(user.features, uents, by="user_id")
```
We can now cluster the users,
```{r usrclus}
user.ftr.sc <- scale(user.features[,-1])
user.ftr.sc <- data.frame(user_id=user.features$user_id, user.ftr.sc)
user.ftr.dis <- dist(user.ftr.sc[,-1], method="euclidean")
user.ftr.mat <- as.matrix(user.ftr.sc[,-1])
hc <- hclust(user.ftr.dis , method="ward.D2")
plot(hc)
#heatmap(user.ftr.mat)
user.ftr.sc <- user.ftr.sc[ order(cutree(hc, h=20)),]
ufm.m <- melt(user.ftr.sc[, -2], id.vars = "user_id")
p <- ggplot(ufm.m, aes(variable, user_id)) + 
      geom_tile(aes(fill=sign(value)*log10(abs(value))), color="white") +
       scale_fill_gradient(low="yellow", high="red")
print(p)
```
