---
title: "Bike sharing from Velopass"
output: html_document
---
#Introduction
We analyze the data from Velopass
#Data
The data in the csvs is in six files, that we load 
```{r data6files}
trans.2010.1 <- read.csv("transactions_page1_2010.csv", sep=";", stringsAsFactors=FALSE)
trans.2010.2 <- read.csv("transactions_page2_2010.csv", sep=";", stringsAsFactors=FALSE)
trans.2010.3 <- read.csv("transactions_page3_2010.csv", sep=";", stringsAsFactors=FALSE)
trans.2010.4 <- read.csv("transactions_page4_2010.csv", sep=";", stringsAsFactors=FALSE)
trans.2010.5 <- read.csv("transactions_page5_2010.csv", sep=";", stringsAsFactors=FALSE)
trans.2010.6 <- read.csv("transactions_page6_2010.csv", sep=";", stringsAsFactors=FALSE)
```
Do all these 6 files have the same dimensions?
```{r data6dim}
dim(trans.2010.1)
dim(trans.2010.2)
dim(trans.2010.3)
dim(trans.2010.4)
dim(trans.2010.5)
dim(trans.2010.6)
```
Page 2 and 4 have 6 columns! Whats the difference? These two files have an extra column *X*, that contains,
```{r diff6pages}
unique(trans.2010.2$X)
```
I will have to ask what these values mean. For now I will ignore the variable *X* and proceed to collate the other columns into one data frame,
```{r collateTransactions}
trans.2010.pages <- list(trans.2010.1, trans.2010.2[, 1:5], trans.2010.3, 
                         trans.2010.4[, 1:5], trans.2010.5, trans.2010.6)
trans.2010 <- do.call("rbind", trans.2010.pages)
```



#Reciprocity
How many from a station, and how many to?
```{r tofromstation}
tofromMat <- table(trans.2010$from_station, trans.2010$to_station)
toDF <- rowSums(tofromMat)
fromDF <- colSums(tofromMat)
